Windowed least-squares model reduction for dynamical systems

### 2019-10-17 main.tex

\begin{appendices}
\section{Proper orthogonal decomposition}
Algorithm~\ref{alg:pod} presents the algorithm for computing the trial basis via proper orthogonal decomposition.
\begin{algorithm}
\caption{Algorithm for generating POD Basis.}
\label{alg:pod}
Input: Number of time-steps between snapshots $N_{\text{skip}}$; intercept $\stateInterceptArg{}$; basis dimension $K$ \; 
%\newline
Output: POD Basis $\basisspace \in \mathbb{V}_{K}(\RR{N})$ \;
%\newline
Steps:
\begin{enumerate}
    \item Solve FOM O$\Delta$E and collect solutions into snapshot matrix
$$\mathbf{S}(N_{\text{skip}}) \defeq \begin{bmatrix} \stateFOMDiscreteArg{0} - \stateInterceptArg{n} & \stateFOMDiscreteArg{N_{\text{skip}}} - \stateInterceptArg{n} & \cdots & \stateFOMDiscreteArg{\text{floor}(N_t/N_{\text{skip}})N_{\text{skip}}} - \stateInterceptArg{n} \end{bmatrix}$$
    \item Compute the thin singular value decomposition, $$\mathbf{S} (N_{\text{skip}}) = \mathbf{U \Sigma Z}^T,$$
    where $\mathbf{U} \equiv \begin{bmatrix} \mathbf{u}_1 & \cdots & \mathbf{u}_{\text{floor}(N_t/N_{\text{skip}})}\end{bmatrix}$
    \item Truncate left singular vectors and to form the basis, $\basisspace \equiv \begin{bmatrix} \mathbf{u}_1 & \cdots & \mathbf{u}_K \end{bmatrix}$
\end{enumerate}

@@

\begin{algorithm}
\caption{Algorithm for generating the sampling matrix through q-sampling.}
\label{alg:qdeim}
Input: Number of time-steps between snapshots $N_{\text{skip}}$, number of primal sampling points, $n_s$ \; 
%\newline
Output: Weighting matrix $\stweightingMatArg{} \equiv \stweightingMatOneTArg{} \stweightingMatOneArg{} \in \{0,1\}^{N \times N}$ \;
%\newline
Steps:
\begin{enumerate}
    \item Solve FOM O$\Delta$E and collect velocity snapshots 
$$\mathbf{F}(N_{\text{skip}}) \defeq \begin{bmatrix} \velocity (\stateFOMDiscreteArg{0})  & \velocity( \stateFOMDiscreteArg{N_{\text{skip}}}) & \cdots & \velocity( \stateFOMDiscreteArg{\text{floor}(N_t/N_{\text{skip}})N_{\text{skip}}} ) \end{bmatrix}$$
    \item Compute the thin singular value decomposition, $$\mathbf{F} (N_{\text{skip}}) = \mathbf{U \Sigma Z}^T,$$
    where $\mathbf{U} \equiv \begin{bmatrix} \mathbf{u}_1 & \cdots & \mathbf{u}_{\text{floor}(N_t/N_{\text{skip}})}\end{bmatrix}$
    \item Compute the QR factorization of  $\mathbf{U}^T$ with column pivoting,
    \begin{equation*}
        \mathbf{U}^T \mathbf{P}^* = \mathbf{QR}
    \end{equation*}
    with $\mathbf{P}^* \equiv \begin{bmatrix} \mathbf{p}_1 & \cdots & \mathbf{p}_{\text{floor}(N_t/N_{\text{skip}})} \end{bmatrix}$, $\mathbf{p}_i \in \{0,1\}^N$. 
    \item Select the first $n_s$ columns of $\mathbf{P}^*$ to form the sampling point matrix $\stweightingMatOneTArg{} \in \{0,1\}^{N \times n_s}$. 
    \item Augment the sampling point matrix, $\stweightingMatOneArg{}$, with additional columns such that all unknowns are computed at the mesh cells selected by Step 3. For the second numerical experiment, these additional unknowns correspond to each conserved variable and quadrature point in the selected cells. 
\end{enumerate}

>>>>

\begin{appendices}
\section{Proper orthogonal decomposition}
Algorithm~\ref{alg:pod} presents the algorithm for computing the trial basis via proper orthogonal decomposition.
\begin{algorithm}
\caption{Algorithm for generating POD Basis.}
\label{alg:pod}
\textbf{Input:} Number of time-steps between snapshots $N_{\text{skip}}$; intercept $\stateInterceptArg{}$; basis dimension $K$ \; 
%\newline
\textbf{Output:} POD Basis $\basisspace \in \mathbb{V}_{K}(\RR{N})$ \;
%\newline
Steps:
\begin{enumerate}
    \item Solve FOM O$\Delta$E and collect solutions into snapshot matrix
$$\mathbf{S}(N_{\text{skip}}) \defeq \begin{bmatrix} \stateFOMDiscreteArg{0} - \stateInterceptArg{n} & \stateFOMDiscreteArg{N_{\text{skip}}} - \stateInterceptArg{n} & \cdots & \stateFOMDiscreteArg{\text{floor}(N_t/N_{\text{skip}})N_{\text{skip}}} - \stateInterceptArg{n} \end{bmatrix}$$
    \item Compute the thin singular value decomposition, $$\mathbf{S} (N_{\text{skip}}) = \mathbf{U \Sigma Z}^T,$$
    where $\mathbf{U} \equiv \begin{bmatrix} \mathbf{u}_1 & \cdots & \mathbf{u}_{\text{floor}(N_t/N_{\text{skip}})}\end{bmatrix}$
    \item Truncate left singular vectors and to form the basis, $\basisspace \equiv \begin{bmatrix} \mathbf{u}_1 & \cdots & \mathbf{u}_K \end{bmatrix}$
\end{enumerate}

@@

\begin{algorithm}
\caption{Algorithm for generating the sampling matrix through q-sampling.}
\label{alg:qdeim}
\textbf{Input:} Number of time-steps between snapshots $N_{\text{skip}}$, number of primal sampling points, $n_s$ \; 
%\newline
\textbf{Output:} Weighting matrix $\stweightingMatArg{} \equiv \stweightingMatOneTArg{} \stweightingMatOneArg{} \in \{0,1\}^{N \times N}$ \;
%\newline
Steps:
\begin{enumerate}
    \item Solve FOM O$\Delta$E and collect velocity snapshots 
$$\mathbf{F}(N_{\text{skip}}) \defeq \begin{bmatrix} \velocity (\stateFOMDiscreteArg{0})  & \velocity( \stateFOMDiscreteArg{N_{\text{skip}}}) & \cdots & \velocity( \stateFOMDiscreteArg{\text{floor}(N_t/N_{\text{skip}})N_{\text{skip}}} ) \end{bmatrix}$$
    \item Compute the thin singular value decomposition, $$\mathbf{F} (N_{\text{skip}}) = \mathbf{U \Sigma Z}^T,$$
    where $\mathbf{U} \equiv \begin{bmatrix} \mathbf{u}_1 & \cdots & \mathbf{u}_{\text{floor}(N_t/N_{\text{skip}})}\end{bmatrix}$
    \item Compute the QR factorization of  $\mathbf{U}^T$ with column pivoting,
    \begin{equation*}
        \mathbf{U}^T \mathbf{P}^* = \mathbf{QR}
    \end{equation*}
    with $\mathbf{P}^* \equiv \begin{bmatrix} \mathbf{p}_1 & \cdots & \mathbf{p}_{\text{floor}(N_t/N_{\text{skip}})} \end{bmatrix}$, $\mathbf{p}_i \in \{0,1\}^N$. 
    \item Select the first $n_s$ columns of $\mathbf{P}^*$ to form the sampling point matrix $\stweightingMatOneTArg{} \in \{0,1\}^{N \times n_s}$. 
    \item Augment the sampling point matrix, $\stweightingMatOneArg{}$, with additional columns such that all unknowns are computed at the mesh cells selected by Step 3. For the second numerical experiment, these additional unknowns correspond to each conserved variable and quadrature point in the selected cells. 
\end{enumerate}

### 2019-10-17 section4/spatial_numerical.tex

\begin{remark}
For the limiting case where $\nstepsArg{n} = 1$, $n=1,\ldots,\nslabs$ such that the window size is equivalent to the time step (i.e., $\DeltaSlabArg{n} = \timeWindowArg{n}{1} - \timeWindowArg{n}{0}$), uniform 
quadrature weights are used, a uniform trial space is employed (i.e., $\trialspaceArg{n} = \trialspace, n=1,\ldots,\nslabs$), the weighting matrices are taken to be
$\stweightingMatOneArg{n} = \lspgWeightingArg{}, n=1,\ldots,\nslabs$, and the time instances $\timeWindowArg{n}{1} = t^n$, $n=1,\ldots,\nslabs$, then 
$\approxstateDiscreteArg{n,1}  = \approxstateLSPG^n$, $n=1,\ldots,\nslabs$ and  
\methodAcronym\ with \spatialAcronym\ trial
	subspaces solved via the direct approach recovers the LSPG approach. 
\end{remark} 
\subsubsection{Solution to the least-squares problem through the Gauss--Newton method}
	\KTC{If the dynamics are linear, doesn't this just yield a linear
	least-squares problem that can be solved directly? We need to present this
	more clearly.}
Discretization through linear multistep methods (as well as other techniques) 
results in a least-squares problem. In particular, for the case where the full-order model~\eqref{eq:FOM} is nonlinear, the direct approach yields a nonlinear least-squares problem. 
A variety of algorithms exist for solving nonlinear least-squares problems, including trust region approaches, the Gauss–Newton method, and the Levenberg–Marquardt method. The numerical experiments presented in this work comprise nonlinear systems and are solved via the Gauss--Newton method, and as such we outline this approach here.

@@

\begin{remark}\label{remark:gaussnewton}\textit{(Acceleration of the Gauss--Newton Solve)}\\
	The principal cost of a Gauss--Newton method is often the formation of the Jacobian matrix. A variety of techniques aimed at 
reducing this computational burden exist; Jacobian-free Newton-Krylov
	methods~\cite{jfnk}, Broyden's method~\cite{broyden} (as explored in
	Ref.~\cite{carlberg_thesis}, Appendix A), and frozen Jacobian approximations
	are several such examples. Further, the space--time formulation introduces an extra dimension for parallelization that can be used to accelerate the wall-clock time of the ROM. The investigation of these additional, potentially more efficient, solution algorithms will be a topic of future work. 
\end{remark}
%\subsubsection{Jacobian-Free Implementation}
%The dominant cost of the Gauss Newton algorithm is the computatation of the action of the Jacobian matrix on the trial basis. To alleviate this burden, the following Jacobian-free method is proposed:
\begin{algorithm}
\caption{\spatialAcronym\ trial subspace: algorithm for the direct solution technique with the Gauss--Newton method and a linear multistep method over the $n$th window}
\label{alg:colloc_gn}
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{tolerance, $\epsilon$; initial guess, $\genstateGuessDiscreteArg{n,1}{0},\ldots,\genstateGuessDiscreteArg{n,\nstepsArg{n}}{0}$}%; initial condition $\genstateDiscreteArg{n,0}$}
\Output{Solution to least squares problem, $\genstatecollocMatSlabArg{n}$} 
\textbf{Online Steps}: \\
$\text{converged} \leftarrow \text{false}$ \Comment{Set convergence checker} \\
$k \leftarrow 0$ \Comment{Set counter}\\
$\genstatecollocMatSlabArg{n}_k \leftarrow \unroll(\genstateGuessDiscreteArg{n,1}{0},\ldots,\genstateGuessDiscreteArg{n,\nstepsArg{n}}{0})$ \Comment{Assemble generalized coordinates over window} \\
\While{\text{converged} == \text{false}}
{
%\For{$i=1,\hdots,\nstepsArg{n}$}{
%  \For{$j=1,\hdots,\ncollocArg{n}{i}$}{
%Compute: $\approxstateArgnt{n,i}  =  \basisspace  \genstateDiscreteArg{n,i} + \stateIntercept$ 
%\Comment{Compute state} \\
%Compute: $\velocity(\basisspace \genstateDiscreteArg{n,i} + \stateIntercept ) $ \Comment{Compute velocity at each time-step instance}\\
%Compute: $\residLMSArg{n,i}(\basisspace \genstateDiscreteArg{n,i} + \stateIntercept, \ldots , \basisspace \genstateDiscreteArg{n,i - k^n(i)} + \stateIntercept) $  \Comment{Compute residual} \\
%}
$\mathbf{r} \leftarrow \residLMSSlabArg{n}(\genstatecollocMatSlabArg{n}_k)$ \Comment{Compute weighted residual over window} \\
$\mathbf{J} \leftarrow  
\frac{\partial \residLMSSlabArg{n}}{\partial \genstatecollocMatySlabArg{}}(\genstatecollocMatSlabArg{n}_k) 
$ \Comment{Compute weighted residual-Jacobian over window} \\
%Compute: $[\jacobianSlabArg{n}]^T \jacobianSlabArg{n}$ \Comment{Compute system matrix for the normal equations} \\
%Compute: $[\jacobianSlabArg{n}]^T \residLMSSlabArg{n}(\genstatecollocMatSlabArg{n}_k)$ \Comment{compute RHS for normal equations} \\
\uIf{ $\norm{ \mathbf{J}^T\mathbf{r}  } \le \epsilon$ }{
{\text{converged} $\leftarrow$ \text{true}}  \Comment{Check and set convergence based on gradient norm} \\
Return: $\genstatecollocMatSlabArg{n} = \genstatecollocMatSlabArg{n}_{k+1} $ \Comment{Return converged solution}\\
}
\Else{
 Compute  $\Delta  \genstatecollocMatSlabArg{n} $ that minimizes $\norm{ \mathbf{J} \Delta \genstatecollocMatSlabArg{n} +\mathbf{r}}^2$ \Comment{Solve the linear least-squares problem} \\
% Compute  $\Delta  \genstatecollocMatSlabArg{n} $ satisfying $ [\mathbf{J}]^T \mathbf{J} \Delta \genstatecollocMatSlabArg{n}=  -\mathbf{J}^T\mathbf{r}$ \Comment{Solve the normal equations} \\
$\genstatecollocMatSlabArg{n}_{k+1} \leftarrow \genstatecollocMatSlabArg{n}_k + \Delta \genstatecollocMatSlabArg{n}$ \Comment{Update guess to the state} \\
}
$k\leftarrow k+1$
}
\end{algorithm}

@@

\subsection{\spatialAcronym\ trial subspaces: indirect solution approach}\label{sec:indirect}
In contrast to the direct approach,
indirect methods ``indirectly" solve the minimization
problem~\eqref{eq:tclsrm} by solving the Euler--Lagrange equations. The
system given by the Euler--Lagrange equations~\eqref{eq:lspg_continuous}--\eqref{eq:lspg_adjoint} comprises a coupled two-point boundary value
problem. A variety of techniques have
been devised to solve two-point boundary value problems of this type. These
techniques include shooting methods, multiple shooting
methods~\cite{multiple_shooting}, and the forward--backward sweep
method~\cite{fbs} (FBSM).  This work explores using the FBSM.

>>>>

\begin{remark}
For the limiting case where $\nstepsArg{n} = 1$, $n=1,\ldots,\nslabs$ such that the window size is equivalent to the time step (i.e., $\DeltaSlabArg{n} = \timeWindowArg{n}{1} - \timeWindowArg{n}{0}$), uniform 
quadrature weights are used, a uniform trial space is employed (i.e., $\trialspaceArg{n} = \trialspace, n=1,\ldots,\nslabs$), the weighting matrices are taken to be
$\stweightingMatOneArg{n} = \lspgWeightingArg{}, n=1,\ldots,\nslabs$, and the time instances $\timeWindowArg{n}{1} = t^n$, $n=1,\ldots,\nslabs$, then 
$\approxstateDiscreteArg{n,1}  = \approxstateLSPG^n$, $n=1,\ldots,\nslabs$ and  
\methodAcronym\ with \spatialAcronym\ trial
	subspaces solved via the direct approach recovers the LSPG approach. 
\end{remark} 
\subsubsection{Solution to the least-squares problem through the Gauss--Newton method}
	\KTC{I recommend rewriting this section with several important things in
	mind. We can discuss.}
	\KTC{If the dynamics are linear, doesn't this just yield a linear
	least-squares problem that can be solved directly? We need to present this
	more clearly.}
Discretization through linear multistep methods (as well as other techniques) 
results in a least-squares problem. In particular, for the case where the full-order model~\eqref{eq:FOM} is nonlinear, the direct approach yields a nonlinear least-squares problem. 
A variety of algorithms exist for solving nonlinear least-squares problems, including trust region approaches, the Gauss–Newton method, and the Levenberg–Marquardt method. The numerical experiments presented in this work comprise nonlinear systems and are solved via the Gauss--Newton method, and as such we outline this approach here.

@@

\begin{remark}\label{remark:gaussnewton}\textit{(Acceleration of the Gauss--Newton Solve)}\\
	The principal cost of a Gauss--Newton method is often the formation of the Jacobian matrix. A variety of techniques aimed at 
reducing this computational burden exist; Jacobian-free Newton-Krylov
	methods~\cite{jfnk}, Broyden's method~\cite{broyden} (as explored in
	Ref.~\cite{carlberg_thesis}, Appendix A), and frozen Jacobian approximations
	are several such examples. Further, the space--time formulation introduces an extra dimension for parallelization that can be used to accelerate the wall-clock time of the ROM. The investigation of these additional, potentially more efficient, solution algorithms will be a topic of future work. 
\end{remark}
%\subsubsection{Jacobian-Free Implementation}
%The dominant cost of the Gauss Newton algorithm is the computatation of the action of the Jacobian matrix on the trial basis. To alleviate this burden, the following Jacobian-free method is proposed:
\begin{algorithm}
\caption{\spatialAcronym\ trial subspace: algorithm for the direct solution technique with the Gauss--Newton method and a linear multistep method over the $n$th window}
\label{alg:colloc_gn}
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{tolerance, $\epsilon$; initial guess, $\genstateGuessDiscreteArg{n,1}{0},\ldots,\genstateGuessDiscreteArg{n,\nstepsArg{n}}{0}$}%; initial condition $\genstateDiscreteArg{n,0}$}
\Output{Solution to least squares problem, $\genstatecollocMatSlabArg{n}$} 
\textbf{Online Steps}: \\
$\text{converged} \leftarrow \text{false}$ \Comment{Set convergence checker} \\
$\genstatecollocMatSlabArg{n}_0 \leftarrow \unroll(\genstateGuessDiscreteArg{n,1}{0},\ldots,\genstateGuessDiscreteArg{n,\nstepsArg{n}}{0})$ \Comment{Assemble generalized coordinates over window} \\
$k \leftarrow 0$ \Comment{Set counter}\\
\While{\text{not converged}}
{
%\For{$i=1,\hdots,\nstepsArg{n}$}{
%  \For{$j=1,\hdots,\ncollocArg{n}{i}$}{
%Compute: $\approxstateArgnt{n,i}  =  \basisspace  \genstateDiscreteArg{n,i} + \stateIntercept$ 
%\Comment{Compute state} \\
%Compute: $\velocity(\basisspace \genstateDiscreteArg{n,i} + \stateIntercept ) $ \Comment{Compute velocity at each time-step instance}\\
%Compute: $\residLMSArg{n,i}(\basisspace \genstateDiscreteArg{n,i} + \stateIntercept, \ldots , \basisspace \genstateDiscreteArg{n,i - k^n(i)} + \stateIntercept) $  \Comment{Compute residual} \\
%}
$\mathbf{r} \leftarrow \residLMSSlabArg{n}(\genstatecollocMatSlabArg{n}_k)$ \Comment{Compute weighted residual over window} \\
$\mathbf{J} \leftarrow  
\frac{\partial \residLMSSlabArg{n}}{\partial \genstatecollocMatySlabArg{}}(\genstatecollocMatSlabArg{n}_k) 
$ \Comment{Compute weighted residual-Jacobian over window} \\
%Compute: $[\jacobianSlabArg{n}]^T \jacobianSlabArg{n}$ \Comment{Compute system matrix for the normal equations} \\
%Compute: $[\jacobianSlabArg{n}]^T \residLMSSlabArg{n}(\genstatecollocMatSlabArg{n}_k)$ \Comment{compute RHS for normal equations} \\
\uIf{ $\norm{ \mathbf{J}^T\mathbf{r}  } \le \epsilon$ }{
{\text{converged} $\leftarrow$ \text{true}}  \Comment{Check and set convergence based on gradient norm} \\
Return: $\genstatecollocMatSlabArg{n} = \genstatecollocMatSlabArg{n}_{k+1} $ \Comment{Return converged solution}\\
}
\Else{
 Compute  $\Delta  \genstatecollocMatSlabArg{n} $ that minimizes $\norm{ \mathbf{J} \Delta \genstatecollocMatSlabArg{n} +\mathbf{r}}^2$ \Comment{Solve the linear least-squares problem} \\
% Compute  $\Delta  \genstatecollocMatSlabArg{n} $ satisfying $ [\mathbf{J}]^T \mathbf{J} \Delta \genstatecollocMatSlabArg{n}=  -\mathbf{J}^T\mathbf{r}$ \Comment{Solve the normal equations} \\
$\genstatecollocMatSlabArg{n}_{k+1} \leftarrow \genstatecollocMatSlabArg{n}_k + \Delta \genstatecollocMatSlabArg{n}$ \Comment{Update guess to the state} \\
}
$k\leftarrow k+1$
}
\end{algorithm}

@@

\subsection{\spatialAcronym\ trial subspaces: indirect solution approach}\label{sec:indirect}
\KTC{This is a fantastic insight and a really great contribution!}
In contrast to the direct approach,
indirect methods ``indirectly" solve the minimization
problem~\eqref{eq:tclsrm} by solving the Euler--Lagrange
equations~\eqref{eq:lspg_continuous}--\eqref{eq:lspg_adjoint} associated with
stationarity. The
system associated with the Euler--Lagrange equations comprises a coupled two-point boundary value
problem. Several techniques have
been devised to solve such problems, 
including shooting methods, multiple shooting
methods~\cite{multiple_shooting}, and the forward--backward sweep
method~\cite{fbs} (FBSM).  This work explores using the FBSM.
